---
title: "BrookHemphill_A05_DataVisualization.Rmd"
author: "Brook Hemphill"
date: "Spring 2024"
output:
  pdf_document: default
  word_document: default
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options:
  chunk_output_type: console
---
\listoffigures
## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 



2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
install.packages('formatR')
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=50), tidy=TRUE)

#1 
getwd()
library(cowplot); library(tidyverse); library(lubridate); library(here)
here()

PeterPaul.chem.nutrients_processed <- 
  read.csv(here("Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"),
           stringsAsFactors = T)
Niwot_Ridge_Litter_processed <-
  read.csv(here("Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"),
           stringsAsFactors = T)

#2 

PeterPaul.chem.nutrients_processed$sampledate <- as.Date(PeterPaul.chem.nutrients_processed$sampledate)
#PeterPaul.chem.nutrients_processed$month <- as.Date(PeterPaul.chem.nutrients_processed$month)
Niwot_Ridge_Litter_processed$collectDate <- as.Date(Niwot_Ridge_Litter_processed$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3
 
mytheme <- theme_minimal(base_size = 14) +
  theme(axis.text = element_text(color = "black", size = 10), 
        legend.position = "top",
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 12),
        plot.title = element_text(hjust = 0.5),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10))

mytheme2 <- theme_minimal(base_size = 14) +
  theme(axis.text = element_text(color = "black", size = 10), 
        legend.position = "top",
        plot.title = element_text(hjust = 0.5),
        axis.title.x = element_blank())
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 

PvPO4 <- ggplot(PeterPaul.chem.nutrients_processed, aes(
  x = po4, y = tp_ug, color = lakename, position = "dodge", shape = lakename)) + 
  geom_point(alpha = 0.8) +
  xlim(0, 35) +
  geom_smooth(method = 'lm', se = FALSE, size = 1) +
  scale_color_manual(values = c("darkblue", "lightblue")) +
  labs(title = "Total Phosphorus vs Phosphate\n", y = "Total Phosphorus (ug)", x = "Phospate") +
  mytheme


print(PvPO4)
  

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.
\newpage

```{r Create boxplots, fig.height=3, fig.width=5, fig.cap= "Figure 1"}
#5 
# months in integers to factors 
PeterPaul.chem.nutrients_processed$month <- factor(PeterPaul.chem.nutrients_processed$month, levels = 1:12, labels = month.abb)

NTL_LTER_temperature <- ggplot(PeterPaul.chem.nutrients_processed, aes(
  x = month, y = temperature_C)) + geom_boxplot(aes(color = lakename), alpha = .8) +
   scale_color_manual(values = c("darkblue", "lightblue")) +
  labs(title = "Temperature each Month\n", y = "Temperature (C)", x = "Month", color = "Lake Name") +
  mytheme2

NTL_LTER_TP <- ggplot(PeterPaul.chem.nutrients_processed, aes(
  x = month, y = tp_ug)) + geom_boxplot(aes(color = lakename), alpha = .8, show.legend = FALSE) +
   scale_color_manual(values = c("darkblue", "lightblue")) +
  labs(title = "Total Phosphorus each Month\n", y = "Total Phosphorus", x = "Month", color = "Lake Name") +
  theme(legend.position = "none", axis.title.x = element_blank()) +
  mytheme

NTL_LTER_TN <- ggplot(PeterPaul.chem.nutrients_processed, aes(x = month, y = tn_ug)) +
  geom_boxplot(aes(color = lakename), alpha = 0.8, show.legend = FALSE) +
  scale_color_manual(values = c("darkblue", "lightblue")) +
  labs(title = "Total Nitrogen each Month\n", y = "Total Nitrogen (ug)", x = "Month", color = "Lake Name") +
  mytheme2

print(NTL_LTER_temperature)
print(NTL_LTER_TP)
print(NTL_LTER_TN)

plot_grid(NTL_LTER_temperature, NTL_LTER_TN, NTL_LTER_TP, nrow = 3, align = 'h', rel_heights = c(1, 1, 1))
```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: Over seasons and between lakes, there is generally more variability in the total phorphorus and nitrogren by month in Peter Lake than in Paul lake. As temperature rises, total P and nitrogen also slightly increase and then decrease as the temperature decreases, however, that trend is slightly harder to see without data from October, Novemeber, and Decemeber.  

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter, fig.height=8, fig.width=8}
#6

needles_subset <- Niwot_Ridge_Litter_processed %>%
  filter(functionalGroup == "Needles")

# Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic
needles_plot <- ggplot(needles_subset, aes(x = collectDate, y = dryMass, color = nlcdClass, shape = nlcdClass)) +
  geom_point(alpha = .5, size = 3) +
 scale_color_manual(values = c("darkgreen", "lightgreen", "darkblue")) +
  labs(title = "Needle Litter Dry Mass by Year\n", y = "Dry Mass", x = "Collection Date", color = "nlcdClass") +
  ylim(0, 50)+ mytheme

print(needles_plot)

#7
needles_plot2 <- ggplot(needles_subset, aes(x = collectDate, y = dryMass, color = nlcdClass, shape = nlcdClass)) +
  geom_point(alpha = .8, size = 3) +
 scale_color_manual(values = c("darkgreen", "lightgreen", "darkblue")) +
  labs(title = "Needle Litter Dry Mass by Year\n", y = "Dry Mass", x = "Collection Date", color = "nlcdClass") +
  ylim(0, 50)+facet_wrap(vars(nlcdClass), nrow = 3) + mytheme

print(needles_plot2)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: I think it depends on what you are looking for, however, for general purposes 6 is more visually appealing, and the information is condensed/comparable, which makes observing trends easier. 